\documentclass[9pt,twocolumn,twoside]{pnas-new}

\usepackage{gensymb}
\usepackage{siunitx}
\usepackage[switch]{lineno}

% new commands
% q value
\newcommand{\qval}[1]{$q<10^{-#1}$}

% species names
\newcommand{\cel}{\emph{C.~elegans}}
\newcommand{\dicty}{\emph{D.~discoideum}}
\newcommand{\ecol}{\emph{E.~coli}}

% gene names
\newcommand{\gene}[1]{\mbox{\emph{#1}}}
% \newcommand{\gene}[1]{\emph{#1}} # for MS word typesetting

\newcommand{\nlp}{\gene{nlp-31}}
\newcommand{\ftna}{\gene{ftn-1}}
\newcommand{\ftnb}{\gene{ftn-2}}
\newcommand{\cysl}{\gene{cysl-1}}
\newcommand{\nog}{\gene{nog-1}}
\newcommand{\nhr}{\gene{nhr-57}}
\newcommand{\lam}{\gene{lam-3}}
\newcommand{\fog}{\gene{fog-2(lf)}}
\newcommand{\egl}{\gene{egl-9(lf)}}
\newcommand{\rhy}{\gene{rhy-1(lf)}}
\newcommand{\vhl}{\gene{vhl-1(lf)}}
\newcommand{\eglvhl}{\gene{egl-9(lf); vhl-1(lf)}}
\newcommand{\eglhif}{\gene{egl-9(lf) hif-1(lf)}}
\newcommand{\hif}{\gene{hif-1(lf)}}

% protein names
\newcommand{\eglp}{EGL-9}
\newcommand{\rhyp}{RHY-1}
\newcommand{\nogp}{NOG-1}
\newcommand{\vhlp}{VHL-1}
\newcommand{\hifp}{HIF-1}
\newcommand{\fogp}{FOG-2}
\newcommand{\nhrp}{NHR-57}
\newcommand{\lamp}{LAM-3}
\newcommand{\cyslp}{CYSL-1}

% DE genes numbers:
\newcommand{\egln}{2,549}
\newcommand{\rhyn}{3,005}
\newcommand{\vhln}{1,275}
\newcommand{\eglvhln}{3,654}
\newcommand{\hifn}{1,075}
\newcommand{\eglhifn}{744}
\newcommand{\fogn}{2,840}
\newcommand{\total}{7,609}

% downstream targets
\newcommand{\egltargets}{4}
\newcommand{\rhytargets}{0}
\newcommand{\vhltargets}{71} % 72 minus vhl-1 (IDed due to deletion)
\newcommand{\hiftargets}{264}
\newcommand{\hifohtargets}{56}

% website commands
\newcommand{\website}{
            \url{https://wormlabcaltech.github.io/mprsq/}
            }
\newcommand{\webref}{
\href{https://wormlabcaltech.github.io/mprsq/}{website}}

\templatetype{pnasresearcharticle} % Choose template

\title{Reconstructing a metazoan genetic pathway with transcriptome-wide epistasis
       measurements}

\author[a, b, 1]{David Angeles-Albores}
\author[a, b, c, 1]{Carmie Puckett Robinson}
\author[a]{Brian A. Williams}
\author[a]{Barbara J. Wold}
\author[a, b]{Paul W. Sternberg}

\affil[a]{Division of Biology and Biological Engineering, Caltech, Pasadena, CA,
          91125, USA}
\affil[b]{Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA}
\affil[c]{Department of Neurology, Keck School of Medicine, University of
          Southern California, Los Angeles, California, 90033, USA}

% Please give the surname of the lead author for the running footer
\leadauthor{Angeles-Albores}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{
Transcriptome profiling quantitatively measures gene expression genome-wide.
There is widespread interest in using transcriptomic profiles as phenotypes for
epistasis analysis. Though epistasis measurements can be performed using
individual transcripts, this results in many scores that must be interpreted
independently. We developed a statistic that summarizes these measurements,
simplifying analysis. Moreover, epistasis analysis has previously only been
performed on cDNA extracted from single cells. We show that whole-organism
RNA-seq can be used to characterize interactions between genes. With the advent
of genome engineering, mutants can be created easily in many organisms. Thus,
phenotyping is now the rate-limiting step towards reconstructing interaction
networks. Our work potentially represents a solution to this problem because
RNA-seq is sensitive to a variety of genetic perturbations.
}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{DAA, CPR and PWS designed experiments. CPR and BAW performed
all experiments. BJW provided resources. DAA developed computational methods and
performed analysis. DAA, CPR and PWS wrote the paper.}
\authordeclaration{The authors declare no conflict of interest.}
\equalauthors{\textsuperscript{1} DAA  and CPR contributed equally to this work.}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be
                     addressed. E-mail: pws@caltech.edu}

% Keywords are not mandatory, but authors are strongly encouraged to provide
% them. If provided, please include two to five keywords, separated by the pipe
% symbol, e.g:
\keywords{Epistasis $|$ Genetic Interaction $|$ Transcriptome $|$ Hypoxia}

\begin{abstract}
  RNA-seq is commonly used to identify genetic modules that respond to
  perturbations. In single cells, transcriptomes have been used as phenotypes,
  but this concept has not been applied to whole-organism RNA-seq. Also,
  quantifying and interpreting epistatic effects using expression profiles
  remains a challenge. We developed a single coefficient to quantify
  transcriptome-wide epistasis that reflects the underlying interactions and
  which can be interpreted intuitively. To demonstrate our approach, we
  sequenced four single and two double mutants of \emph{Caenorhabditis~elegans}.
  From these mutants, we reconstructed the known hypoxia pathway. In addition,
  we uncovered a class of \hifohtargets{} genes with \gene{hif-1}-dependent
  expression that have opposite changes in expression in mutants of two genes
  which cooperate to negatively regulate \hifp{} abundance; however, the double
  mutant of these genes exhibits suppression epistasis. This class violates the
  classical model of HIF-1 regulation, but can be explained by postulating a
  role of hydroxylated HIF-1 in transcriptional control.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\begin{document}

% Optional adjustment to line up main text (after abstract) of first page with
% line numbers, when using both lineno and twocolumn options. You should only
% change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\linenumbers{}

\begin{table}[tbhp]
  \centering
  \begin{tabular}{lr}
    \toprule{}
    Genotype & Differentially Expressed Genes\\
    \midrule{}\egl{} & \egln{}\\
    \rhy{} & \rhyn{}\\
    \vhl{} & \vhln{}\\
    \hif{} & \hifn{}\\
    \eglvhl{} & \eglvhln{}\\
    \eglhif{} & \eglhifn{}\\
    \fog{} & \fogn{}\\
    \bottomrule{}
  \end{tabular}
  \caption{Number of differentially expressed genes in each mutant strain with
  respect to the wild type (N2).}
\label{tab:genes}
\end{table}

\end{document}
